Distributional Trees and Forests for Circular Data

### 2019-09-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small revisions on methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2765 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 38da6d9010bf8be5be44b393f8bcea77bddff7b8

\subsection{Circular Distributional Tree}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a better 
fitting overall model.
This is the general idea of regression trees which are combined with distributional regression
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
In the presented distributional trees a full distributional model is specified in each node of the 
tree. For this purpose a distribution family has to be chosen in advance. Hence, selecting the 
von Mises distribution allows for an application of distributional trees to circular data following 
the tree building algorithm as introduced by \cite{Schlosser+Hothorn+Stauffer:2019}.

@@

Once a circular distributional tree model is fit it can be employed to obtain probabilistic predictions
for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
the corresponding subset of learning observations. 
This can also be expressed by employing weights which indicate whether the $i$-th learning observation and the observation $\bm{z}$ belong to the same terminal
node:
\begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
partitioning the covariate space in disjoint subsets. 
In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
von Mises distribution for a given $\bm{z}$ is obtained by
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
\end{equation}

@@

\subsection{Circular Distributional Forest}
\label{sec:circforest}
A natural extension of circular distributional trees are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions
by adaptive local likelihood estimation of the distributional parameters
\citep{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on how often $\bm{z}$ is assigned to the same terminal node as
each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Hence, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

In that way a specific set of weights can be calculated for each observation yielding
its specific parameter estimates
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i^{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the unweighted
full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$ corresponds to the abrupt
splits from the tree.
Thus circular distributional forests can capture both, smooth and abrupt changes, while
covariates and possible interactions are selected automatically allowing for an easy to
apply methodology to fully specify the von Mises distribution for circular data.

@@

\subsection{Reference Models}
Next to circular distributional trees and forests three other models have been applied and their 
performances have been compared based on a circular version of the continuous ranked probability 
score (CRPS).

>>>>

\subsection{Distributional Trees for Circular Data}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a better 
fitting overall model.
This is the general idea of regression trees which are combined with distributional regression
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
In the presented distributional trees a full distributional model is specified in each node of the 
tree. For this purpose a distribution family has to be chosen in advance. Hence, selecting the 
von Mises distribution allows for an application of distributional trees to circular data following 
the tree building algorithm as introduced by \cite{Schlosser+Hothorn+Stauffer:2019}.

@@

Once a distributional tree model is fit it can be employed to obtain probabilistic predictions
for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
the corresponding subset of learning observations. 
This can also be expressed by employing weights which indicate whether the $i$-th learning observation and the observation $\bm{z}$ belong to the same terminal
node:
\begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
partitioning the covariate space in disjoint subsets. 
In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
von Mises distribution for a given $\bm{z}$ is obtained by
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
\end{equation}

@@

\subsection{Distributional Forests for Circular Data}
\label{sec:circforest}
A natural extension of distributional trees for circular data are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions
by adaptive local likelihood estimation of the distributional parameters
\citep{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Hence, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

In that way a specific set of weights can be calculated for each observation yielding
its specific parameter estimates for the von Mises distribution
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i^{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the unweighted
full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$ corresponds to the abrupt
splits from the tree.
Thus distributional forests for circular data can capture both, smooth and abrupt changes, while
covariates and possible interactions are selected automatically allowing for an easy to
apply methodology to fully specify the von Mises distribution for circular data.

@@

\subsection{Reference Models}
Next to distributional trees for circular data and forests three other models have been applied and their 
performances have been compared based on a circular version of the continuous ranked probability 
score (CRPS).
